---
title: "class009"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## EDA of PDB general CSV data
```{r}
PDB <- read.csv("rbcs_data_exp.csv", row.names = 1)
head(PDB)

# %s of structures in the PDB 
Q1 <- (sum(PDB$X.ray)+sum(PDB$EM))/sum(PDB$Total)
Q2 <- (sum(PDB[1,7]))/sum(PDB$Total)
```

#Q1. 92.6% of structures in the PDB are solved by X-Ray and EM. 

#Q2. 87.3% of structures in the BDP are proteins.

#Q3. There seem to be 1868 HIV-1 protease structures in the current PDB.


## Visualizing the HIV-1 Protease Structure

#Q4. I believe that we see water as just one atom per molecule because the hydrogens are implied as their exact location cannot be precisely determined (minimum resolving size exceeds ~1A bond length).

#Q5. It isn't very clear as to where the binding site is at this point, but I identified OH308 as a possibility. 

#Q6. I think the low B-value structures between the alpha helices and beta folds are the most likely to only take on that position and form those folds when the protein is dimerized. 


## Intro to Bio3D in R & Comparative Structure Analysis of Adenylate Kinase

```{r}
# Organize Workspace
library(bio3d)

#Access online PDB file
pdb <- read.pdb("1hsg")

print(pdb)

```

#Q7: How many amino acid residues are there in this pdb object? -> 198

#Q8: Name one of the two non-protein residues? -> HOH

#Q9: How many protein chains are in this structure? -> 2

#Q10. Msa is found only on BioConductor and not CRAN.

#Q11. Bio3D-view is not found on BioConductor or CRAN.

#Q12. Functions from the devtools package can be used to install packages from GitHub and BitBucket? -> TRUE

```{r}
# Getting Sequence for PDB Entry
aa <- get.seq("1ake_A")
aa
```

#Q13. There are 214 amino acids in this sequence.

```{r}
#Blast search the loaded sequence
b <- blast.pdb(aa)

#Plot blast results
hits <- plot.blast(b)

#Download PDB files from hit list
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)

# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")

# Vector containing PDB codes for figure axis
ids <- basename.pdb(pdbs$id)

# Draw schematic alignment
plot(pdbs, labels=ids)
#white = missing
#grey = aligned
```


## PCA

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)

# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```


## Normal Mode Analysis

```{r}
# NMA of all structures
modes <- nma(pdbs)

plot(modes, pdbs, col=grps.rd)

```

#Q14. The black and colored lines are different in some places, but the same in others. I assume that the places where the black and color lines overlap is where the structure is conserved between the two conformations, and where they are divergent denotes the places where the two conformations differ the most (residues 40-60 and 125-150).