---
title: "class12"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Differential Expression Analysis

```{r package import}
#Load Packages
library(DESeq2)
library(ggplot2)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(pathview)
library(gage)
library(gageData)
```

## Data Formatting
```{r data import}
#Load Data Files
colData = read.csv("GSE37704_metadata.csv") #metadata file
countData = read.csv("GSE37704_featurecounts.csv", row.names=1) #counts file
```

Length column in countData is not relevant for further processing...
Get colData and countData to match in format [DO NOT REPEAT - will mess up code]
```{r data formatting}
#colData <- t(colData) #DO NOT transpose data to match countData <-> inverse x & y are OK!!!
countData <- countData[,-1] #Remove column `length` because colData doesn't have reference to it
```

Log transformations will not work on zero values, and genes with zeros across all cells
```{r zero removal}
#Remove zero values from countData
countData <- countData[rowSums(countData) != 0, ]
```

```{r data setup}
#Put Data together in DESeq2 format
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
```

## Running DESeq2 
```{r deseq2 analysis}
#Run the DESeq2 pipeline
dds = DESeq(dds)
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))

#Visualize results
summary(res) #4349 genes up-regulated (27%) #4396 genes down-regulated (28%)
```

## Volcano Plot

To visualize the results, we can make a volcano plot. We will color the points based on their relevance: grey = default, red = only fold change > 2, blue = adj p-value < 0.01 && fold change > 2. The colorings proceed in the order from least stringent requirements to most stringent requirements because each step colors over points that already have a color. 
```{r volcano plot}
#Base plot code
#plot( res$log2FoldChange, -log(res$padj) )

# Make a color vector for all genes; same length as the number of results points
volcano.colors <- rep("gray", nrow(res) ) 
volcano.colors[ abs(res$log2FoldChange) > 2 ] <- "red" #color corresponding points blue
volcano.colors[ (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 ) ] <- "blue" #color corresponding points red (red is last bc it has the most stringent requirement) - THESE ARE THE POINTS WE ARE LOOKING FOR

#Replot the plot
plot( res$log2FoldChange, -log(res$padj), col=volcano.colors, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

## Adding Gene Annotation 

KEGG Pathway Analysis uses Entrez IDs, not Ensembl IDs, so we need to add a traslation column to res (DESeq2 results file)
```{r}
#Add Entrez Gene IDs as a column in res
res$Entrez <- mapIds(org.Hs.eg.db,
                     keys = row.names(res),
                     keytype = "ENSEMBL",
                     column = "ENTREZID",
                     multiVals = "first")

#Add Gene Symbols as a column in res
res$Symbol <- mapIds(org.Hs.eg.db,
                     keys = row.names(res),
                     keytype = "ENSEMBL",
                     column = "SYMBOL",
                     multiVals = "first")

#Add Gene Names as a column in res
res$Name <- mapIds(org.Hs.eg.db,
                     keys = row.names(res),
                     keytype = "ENSEMBL",
                     column = "GENENAME",
                     multiVals = "first")
```

```{r}
# Reorder results by adjusted p-value and SAVE to a CSV file.
write.csv(res[order(res$pvalue),], file="DESeq_results.csv")
```

# Section 2. Pathway Analysis

Here we are going to use the gage package for pathway analysis. Once we have a list of enriched pathways, we're going to use the pathview package to draw pathway diagrams, shading the molecules in the pathway by their degree of up/down-regulation.

## KEGG Pathways
```{r load and setup KEGG databases}
data(kegg.sets.hs)
data(sigmet.idx.hs)

# Select for signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]
```

The main `gage()` function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.
```{r format data for GAGE input}
#Make the data vector first 
gage.folds <- res$log2FoldChange

#Assign names to vector values
names(gage.folds) <- res$Entrez
```

```{r get the GAGE results}
#Run the gage pathway analysis pipeline
keggres = gage(gage.folds, gsets=kegg.sets.hs, same.dir=TRUE)
```

```{r KEGG visualization}
#Saves as a PNG
pathview(gene.data=gage.folds, pathway.id="hsa04110")
#Saves as a PDF
pathview(gene.data=gage.folds, pathway.id="hsa04110", kegg.native=FALSE)
```


## Top Five
```{r KEGG top 5}
## Focus on top 5 up-regulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids

#Visualize with pathview()
pathview(gene.data=gage.folds, pathway.id=keggresids, species="hsa")
```


## Bottom Five
```{r KEGG bottom 5}
## Focus on top 5 up-regulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$less)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids

#Visualize with pathview()
pathview(gene.data=gage.folds, pathway.id=keggresids, species="hsa")
```

> Q5. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?
^ Done. (preceeding code chunk)

# Section 3. Gene Ontology (GO)
We can also do a similar procedure with gene ontology. Similar to above, go.sets.hs has all GO terms. go.subs.hs is a named list containing indexes for the BP, CC, and MF ontologies. Ontologies will tell us what specific processes are being dysregulated.

```{r gene ontology}
data(go.sets.hs)
data(go.subs.hs)

#Select for biological processes only
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(gage.folds, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```


# Section 4. Reactome Analysis
Reactome is database consisting of biological molecules and their relation to pathways and processes.

```{r over-representation enrichment analysis}
sig.genes <- res[res$padj <= 0.05 & !is.na(res$padj), "Symbol"]
print(paste("Total number of significant genes:", length(sig.genes)))

#Save data
write.table(sig.genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

> Q6. The Endosomal/Vacuolar pathway has the most significant Entities p-value. No, the most significant pathways listed do not match previous KEGG results. There seems to not be anything in the code selecting for the highest/lowest expressing results from the gene ontology database. 


```{r}
#I am getting confused, so need to reorganize my data
#First, I make my working data table
working.table <- data.frame(countData[ colData$condition == "control_sirna" ], countData[ colData$condition == "hoxa1_kd" ])
```



